---
title: "Density-Distribution-Quantile-Simulation"
author: "Jette Steinbach"
date: "10 11 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This package includes some functions that compute the basic properties of phase-type distributions, i.e. the density, distribution and quantile function as well as simulations.

### The density

#### - in the discrete case
In the following, let $\tau= \inf\{t\geq 1 \mid X_t = p+1\}$ be discrete phase-type distributed with initial distribution $\pi$ and sub-transition probability matrix $T$, $\tau \sim DPH_{p}(\pi,T)$.\
Then the density for $\tau$ is given by
\begin{align*}
\mathbb{P}(\tau= n) &= \sum_{i=1}^{p} \mathbb{P}(\tau=n \mid X_{n-1}=i)\mathbb{P}(X_{n-1}=i)\\
&= \sum_{i=1}^{p} \mathbb{P}(\tau=n \mid X_{n-1}=i) \sum_{j=1}^{p} \mathbb{P}(X_{n-1}=i \mid X_0 =j ) \mathbb{P}(X_0=j)\\
&= \sum_{i=1}^{p} \sum_{j=1}^{p} \mathbb{P}(\tau=n \mid X_{n-1}=i)  \mathbb{P}(X_{n-1}=i \mid X_0 =j ) \mathbb{P}(X_0=j)\\
\end{align*}
for $n \geq 1$, where the first and second equality are due to the law of total probability. Now we observe that the probability $\mathbb{P}(X_0=j)$ for the Markov jump process $(X_t)_{t\geq 0}$ to start in $j$ is equal to the $j$'th entry of the initial distribution. Furthermore, the probability $\mathbb{P}(\tau=n \mid X_{n-1}=i)$ for the process to be absorbed at time $n$, given that the process is in state $i$ at time $n-1$ is given by the $i$'th entry of the exit probability vector $t = \boldsymbol{e} - T \boldsymbol{e}$. Finally, the probability $\mathbb{P}(X_{n-1}=i \mid X_0 =j )$ for the process to be in state $i$ at time $n-1$ given that the process started in state $j$ (or equivalently the probability of going from state $j$ to state $i$ in time $n-1$) is given by the $(j,i)$'th element of sub-transition probability matrix $T^{n-1}$. Hence, we have that  
\begin{align*}
\mathbb{P}(\tau= n) &= \sum_{i=1}^{p} \sum_{j=1}^{p} \mathbb{P}(\tau=n \mid X_{n-1}=i)  \mathbb{P}(X_{n-1}=i \mid X_0 =j ) \mathbb{P}(X_0=j)\\
&= \sum_{i=1}^{p} \sum_{j=1}^{p} t_i (T^{n-1} )_{ji} \pi_j\\
&= \sum_{i=1}^{p} \sum_{j=1}^{p} \pi_j (T^{n-1} )_{ji} t_i\\
&= \pi T^{n-1} t.
\end{align*}

We also need to take possible defect sizes into account. So let $\tau \sim DPH_{p}(\pi,T)$ with defect size $\pi_{p+1}$, i.e.
\begin{equation*}
\tau \sim \begin{cases}
1 & \text{ with probability $\pi_{p+1}$} \\
DPH_{p}(\pi,T) & \text{ with probability $1- \pi_{p+1}$} .
\end{cases}
\end{equation*}
Then we have that 
\begin{align*}
\mathbb{P}(\tau= n) &= \mathbb{P}(\tau \cdot \mathbb{1}_{\{\tau \geq 1\}}= n)\\
&= \mathbb{P}(\tau(\mathbb{1}_{\{\tau = 1\}} + \mathbb{1}_{\{\tau > 1\}}) = n)\\
&= \mathbb{P}(\tau\cdot \mathbb{1}_{\{\tau = 1\}} = n) + \mathbb{P}(\tau\cdot \mathbb{1}_{\{\tau > 1\}} = n)
\end{align*}
Due to the definition of $\tau$, we know that when $\tau > 1$, $\tau \sim DPH_{p}(\pi,T)$. Hence, from the previous computations
\begin{equation*}
\mathbb{P}(\tau\cdot \mathbb{1}_{\{\tau > 1\}} = n) = \pi T^{n-1} t \quad \text{ for } n\geq 1.
\end{equation*}
On the other hand, 
\begin{equation*}
\mathbb{P}(\tau \cdot \mathbb{1}_{\{\tau = 1\}} = n) = \mathbb{P}(1 \cdot \mathbb{1}_{\{\tau = 1\}} = n) = \begin{cases}
\mathbb{P}(\tau = 1 )=\pi_{p+1}  & \text{ if } n=1\\
0 & \text{ otherwise}.
\end{cases}
\end{equation*}
This implies that 
\begin{align*}
\mathbb{P}(\tau= n) &= \mathbb{P}(\tau\cdot \mathbb{1}_{\{\tau = 1\}} = n) + \mathbb{P}(\tau\cdot \mathbb{1}_{\{\tau > 1\}} = n)\\
&= \begin{cases}
\pi T^{n-1} t + \pi_{p+1}  & \text{ if } n=1\\
\pi T^{n-1} t & \text{ otherwise}.
\end{cases}
\end{align*}

#### - in the continuous case
Now let $\tau= \inf\{t > 0 \mid X_t = p+1\}$ be continuous phase-type distributed with initial distribution $\pi$ and sub-intensity rate matrix $T$, $\tau \sim PH_{p}(\pi,T)$.\
Then the density for $\tau$ is given by the derivative of the distribution function (see next section)
\begin{align*}
f(u) &= \frac{\partial}{\partial u} F(u)\\
&= \frac{\partial}{\partial u} 1-\pi e^{Tu} \boldsymbol{e} \\
&= -\pi e^{Tu}T \boldsymbol{e}\\
&= \pi e^{Tu}t
\end{align*}
for $u \geq 0$, where we have used that $t = -T \boldsymbol{e}$.

### The distribution function

#### - in the discrete case
Let again $\tau= \inf\{t\geq 1 \mid X_t = p+1\}$ be discrete phase-type distributed with initial distribution $\pi$ and sub-transition probability matrix $T$, $\tau \sim DPH_{p}(\pi,T)$.\
Then the distribution function for $\tau$ is given by
\begin{align*}
F_{\tau}(n) = \mathbb{P}(\tau \leq n) &= 1- \mathbb{P}(\tau > n)\\
&= 1-\sum_{i=1}^{p} \mathbb{P}( X_{n}=i)\\
&= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} \mathbb{P}( X_{n}=i \mid X_0 =j) \mathbb{P}( X_0 =j)
\end{align*}
for $n \geq 1$, where the fourth equality is due to the law of total probability. As before, we observe that the probability $\mathbb{P}(X_0=j)$ for the Markov jump process $(X_t)_{t\geq 0}$ to start in $j$ is equal to the $j$'th entry of the initial distribution, and that the probability $\mathbb{P}(X_{n}=i \mid X_0 =j )$ for the process to be in state $i$ at time $n$ given that the process started in state $j$ (or equivalently the probability of going from state $j$ to state $i$ in time $n$) is given by the $(j,i)$'th element of sub-transition probability matrix $T^{n}$. Hence, we have that  
\begin{align*}
F_{\tau}(n) &= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} \mathbb{P}( X_{n}=i \mid X_0 =j) \mathbb{P}( X_0 =j)\\
&= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} (T^{n})_{ji}\pi_j\\
&= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} \pi_j(T^{n})_{ji}\\
&= 1- \pi T^{n}\boldsymbol{e}.
\end{align*}

#### - in the continuous case
As before, let $\tau= \inf\{t > 0 \mid X_t = p+1\}$ be continuous phase-type distributed with initial distribution $\pi$ and sub-intensity rate matrix $T$, $\tau \sim PH_{p}(\pi,T)$.\
Then the distribution function for $\tau$ is given by
\begin{align*}
 F_{\tau}(u) = \mathbb{P}(\tau \leq u) &= 1- \mathbb{P}(\tau > u)\\
&= 1-\sum_{i=1}^{p} \mathbb{P}( X_{u}=i)\\
&= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} \mathbb{P}( X_{u}=i \mid X_0 =j) \mathbb{P}( X_0 =j)
\end{align*}
for $u \geq 0$, where the fourth equality is due to the law of total probability. As in the discrete case, we observe that the probability $\mathbb{P}(X_0=j)$ for the Markov jump process $(X_t)_{t\geq 0}$ to start in $j$ is equal to the $j$'th entry of the initial distribution, and that the probability $\mathbb{P}(X_{u}=i \mid X_0 =j )$ for the process to be in state $i$ at time $u$ given that the process started in state $j$ (or equivalently the probability of going from state $j$ to state $i$ in time $u$) is given by the $(j,i)$'th element of sub-transition probability matrix $e^{Tu}$. Hence, we have that  
\begin{align*}
F_{\tau}(n) &= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} \mathbb{P}( X_{n}=i \mid X_0 =j) \mathbb{P}( X_0 =j)\\
&= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} (e^{Tu})_{ji}\pi_j\\
&= 1-\sum_{i=1}^{p} \sum_{j=1}^{p} \pi_j(e^{Tu})_{ji}\\
&= 1- \pi e^{Tu}\boldsymbol{e}.
\end{align*}

### The quantile function

In this package, we used the inbuild function `uniroot` to find the quantile for a given probability $p$. Type `help("uniroot")` into your console to find out more about this function.

### Simulations

For an object of type `discphasetype` with initital distribution $\pi$ and sub-intensity rate matrix $T$ the simulation is conducted roughly in the following way. 

* The waiting time $\tau$ is set to be $1$, which corrsponds to immediate absorbtion.
* The initial state $X_1$ is sampled from all states according to the initial distribution $\pi$.
* As long as $X_t$ is differnt from the absorbing state $p+1$, $t\geq1$, we add one to the waiting time $\tau$, i.e. $\tau = \tau +1$, and the current state $X_t$ is sampled among all states $1,...,p+1$ according to the transition probabilites given in $T$ and the exit rates from $t$. 

For an object of type `contphasetype` with initital distribution $\pi$ and sub-transition probability matrix $T$ the simulation is conducted roughly in the same way. 

* The waiting time $\tau$ is set to be $0$, which corrsponds to immediate absorbtion.
* The initial state $X_1$ is sampled from all states according to the initial distribution $\pi$.
* As long as $X_t$ is differnt from the absorbing state $p+1$, $t\geq1$, we add an exponential distributed variable to the waiting time $\tau$, i.e. $\tau = \tau + \lambda$, where $\lambda \sim \text{Exp}(-T[X_t,X_t])$. Furthermore, the current state $X_t$ is sampled among all states $1,...,p+1$ (except for the previous state $X_{t-1}$) according to the rates given in $T$ and the exit rates from $t$.

### Example 5

We want to reproduce Figure 3.4 in John Wakeley (2009): "Coalescent Theory: An Introduction", Roberts and Company Publishers, Colorado, which displays the distributions of $T_{\text{MRCA}}$ and $T_{\text{Total}}$ for $n \in \{2,5,10,20,50,100\}$. The initial distributions and sub-intensity rate matrices are stored in the datasets `T_MRCA` and `T_Total` that are provided in the package.
First, we compute the distribution of $T_{\text{MRCA}}$ 

```{r}

## Defining the vector of quantiles
t.vec <- seq(0,4, by=0.1)

## and the matrix holding all distributions
dist.mat <- matrix(nrow = 6, ncol = length(t.vec))

## Now we can compute the distributions
for(x in t.vec){

  dist.mat[2,which(t.vec ==x)] <- dphasetype(T_MRCA$n5,x)
  dist.mat[3,which(t.vec ==x)] <- dphasetype(T_MRCA$n10,x)
  dist.mat[4,which(t.vec ==x)] <- dphasetype(T_MRCA$n20,x)
  dist.mat[5,which(t.vec ==x)] <- dphasetype(T_MRCA$n50,x)
  dist.mat[6,which(t.vec ==x)] <- dphasetype(T_MRCA$n100,x)
}

## For n=2, the initial distribution is equal to 1 and 
## the sub-intensity rate matrix is T.mat = -1. Hence,
## the distribution is given by
dist.mat[1,] <- exp(-t.vec)

```

Now that we have computed the distribution of $T_{\text{MRCA}}$ for all $n\in \{2,5,10,20,50,100\}$ and $t \in \{0,...,4\}$, we are able to reproduce the first figure
```{r}
plot(t.vec, dist.mat[1,], type = "l", 
     main = expression(paste("The distribution of ", T["MRCA"], 
     " for n=2,5,10,20,50,100")), cex.main = 0.9, xlab = "t", 
     ylab = expression(f[T[MRCA]](x)), xlim = c(0,4), ylim = c(0,1), 
     frame.plot = F)

points(t.vec, dist.mat[2,], type = "l")
points(t.vec, dist.mat[3,], type = "l")
points(t.vec, dist.mat[4,], type = "l")
points(t.vec, dist.mat[5,], type = "l")
points(t.vec, dist.mat[6,], type = "l")

```

Now, we can perform the same calculations for $T_{\text{Total}}$

```{r}

## Defining the vector of quantiles
t.vec <- seq(0,15, by=0.1)

## and the matrix holding all distributions
dist.mat <- matrix(nrow = 6, ncol = length(t.vec))

## Now we can compute the distributions
for(x in t.vec){

  dist.mat[2,which(t.vec ==x)] <- dphasetype(T_Total$n5,x)
  dist.mat[3,which(t.vec ==x)] <- dphasetype(T_Total$n10,x)
  dist.mat[4,which(t.vec ==x)] <- dphasetype(T_Total$n20,x)
  dist.mat[5,which(t.vec ==x)] <- dphasetype(T_Total$n50,x)
  dist.mat[6,which(t.vec ==x)] <- dphasetype(T_Total$n100,x)
}

## For n=2, the initial distribution is equal to 1 and 
## the sub-intensity rate matrix is T.mat = -1/2. Hence,
## the distribution is given by
dist.mat[1,] <- exp(-t.vec/2)/2

```

Now that we have computed the distribution of $T_{\text{Total}}$ for all $n\in \{2,5,10,20,50,100\}$ and $t \in \{0,...,15\}$, we are able to reproduce the second figure
```{r}
plot(t.vec, dist.mat[1,], type = "l", 
     main = expression(paste("The distribution of ", T["Total"], 
     " for n=2,5,10,20,50,100")), cex.main = 0.9, xlab = "t", 
     ylab = expression(f[T[Total]](x)), xlim = c(0,15), ylim = c(0,0.5), 
     frame.plot = F)

points(t.vec, dist.mat[2,], type = "l")
points(t.vec, dist.mat[3,], type = "l")
points(t.vec, dist.mat[4,], type = "l")
points(t.vec, dist.mat[5,], type = "l")
points(t.vec, dist.mat[6,], type = "l")

```

